Untitled

### 2016-12-02 Composition_Ecology.Rmd
Revisions ready to pass back to @iceageecologist.
Generally minor text changes, some changes to the `bib` file.  Filled in values,

Cleaned up the first `if` statement in `checker` in `confounding_table`.

Removed sets of files as well to clean up the repo.

* Anthropocene, climate change, climate disequillibrium, forest inventory and analysis (FIA), fundamental niche, historical ecology, land use, niche shift, Public Land Survey System, realized niche

@@

Contemporary forest inventory data are widely used to understand environmental controls on tree species distributions and to construct models to project forest responses to climate change, but the stability and representativeness of contemporary tree-climate distributions are poorly understood. We show that tree-climate relationships for 15 tree genera in the upper Midwestern US have significantly altered over the last two centuries due to historical land use and climate change. Realized niches have shifted towards higher minimum temperatures, cooler maximum temperatures, and higher rainfall. A new attribution method implicates both historical climate change and land use in these shifts, with the relative importance varying among genera and climate variables. Most climate/land-use interactions are compounding, in which historical land-use reinforces climatic shifts in species distributions toward cooler or wetter distributions. The prevalence of compounding interactions imply that contemporary-based models of species distributions may underestimate species resilience to climate change.

@@

Increasingly, ecologists are turning to the paleohistorical record to test the assumptions underlying ecological models and improve their predictive capacity [@moritz2013future;@maguire2015modeling]. Projections relying only on contemporary observational data may overlook plasticity in species-climate responses [@maiorano2013building], incompletely delimit species fundamental niches [@jackson2000responses], miss responses to past no-analog climates [@veloz2012no], or fail to detect effects of anthropogenic land use, and other historical legacies [@svenning2015influence]. Paleoecological records show that forest responses lag abrupt climate change by a few years to centuries [@williamsinpress;@webb1986vegetation;@ordonez2013realized;@svenning2015influence], suggesting that tree distributions have yet to fully respond to recent climate changes. In principle, mechanistic dynamic vegetation models [*e.g.*, @leiblein2016modelling] can represent the physiological, demographic, and dispersal processes that cause tree populations to lag climate change.  In practice, however, the parameterization of these models is poorly constrained for processes operating at long time scales [@moorcroft2006close;@matthes2016benchmarking].

Historical land use has substantially affected species distributions and vegetation structure [*e.g.*, @gehrig2007tree] and may cause species distributions to incompletely fill available climates, or otherwise be at disequilibrium [@early2014climatic;@sarmento2013impacts;@ivory2016niche]. Anthropogenic land use is structured and is often biased within climate space, hence, species distributions and protected natural areas often represent a biased selection from the available environmental space [@scott2001nature]. Biasing can confound studies that use contemporary distributional data to understand the environmental controls on species distributions and predict species responses to future climate scenarios [@pyke2004habitat].

Interactions between climate and land-use change can be classified into three types: compounding, confounding and counteracting [modified from @pyke2004habitat]. Compounding effects occur when climatic and land-use changes cause species distributions to shift in the same direction along an environmental gradient.  For example, warming has been compounded by land-use conversion at the lowest elevations in the northern Sierra Nevada Mountains of California, both causing upslope distributional shifts for butterfly species [@forister2010compounded]. Confounding effects may cause little overall change in species distributions, with land-use pressure occurring across the species' environmental gradient. For example, land-use conversion has excluded Garry oak (*Quercus garryana*) from the deep-soil sites it formerly occupied [@pellatt2014environmental], but land use has not systematically shifted Garry oak distributions in climate space.  These confounding effects may hinder detection of climatic forcing of historical species distributions. Counteracting effects occur when climate and land use change act in opposite directions, narrowing the available geographic and climatic space available for species. For example, bird species in the Sierra Nevadas are expected to move downslope due to increasing precipitation [@tingley2012push], but this downslope migration is limited by land-use conversion in the lowlands.

@@

`r fig_nums("land_use_clim", "*(a-e) Climate change in the upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions with a LOESS smooth superimposed for illustration. Modern climate (normals indicated by rectangles).  PDSI shows strong coherence with $P_{ann}$ in this region. (f) Patterns of land-use change in the upper Midwestern United States, with the 'Tension Zone', differentiating sub-boreal from southern broadleafed forests superimposed. Land use data from the NLCD [@jin2013comprehensive].*")`

Tree distributions and community composition in the upper Midwestern US have been heavily influenced by land use conversion associated with Euro-American Settlement in the 19^th^ and 20^th^ Centuries [@goring2016preset;@schulte2007homogenization;@hanberry2012comparison;@rhemtulla2009historical], outbreaks of exotic pathogens [@barnes1979stem], climate changes accompanying the Little Ice Age [@hotchkiss2007response] and current anthropogenic warming [@wicci2011climate]. Modern forests show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System (PLS) prior to widespread land-use conversion in the late 1800s [@goring2016preset;@schulte2007homogenization]. Rhemtulla et al. [-@rhemtulla2009historical] estimate that *Pinus strobus* and *Tsuga canadensis* now occupy only 4% of their original coverage in Wisconsin. Wildland ecosystems have been converted to agricultural and urban land use [@rhemtulla2009historical], particularly in historical prairie and savanna ecosystems (`r fig_nums("land_use_clim", display = "cite")`f), where almost 50% of southern deciduous savanna was converted to cropland [@rhemtulla2007regional].  In northern forests, ranges of early successional species have expanded and abundances of late-successional species have been reduced due to widespread logging in the 19^th^ and 20^th^ centuries, and ongoing land management [@schulte2007homogenization;@goring2016preset].  Additionally, the formerly dominant *Ulmus americana* has experienced rangewide population collapses caused by the introduction of exotic pathogens [@barnes1979stem], as did *Castanea dentata*, with other tree species currently threatened (*Tsuga canadensis*, *Fraxinus* spp.) [@barnes1979stem;@schlarbaum1998three;@castello1995pathogens].

Here we demonstrate significant shifts in the climatic distributions of tree genera in the upper Midwestern US over the last two centuries due to historical climate and vegetation change, and we diagnose the relative influence of historical climate change and, land use changes and pathogen outbreaks (for *Ulmus*) on these shifts.  We use gridded historical and contemporary climate observations (PRISM-LT) and estimates of pre-settlement and modern forest vegetation [@goring2016preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwest.  We construct four observed and hypothetical sets of tree-climate relationships, using a 2x2 factorial of pre-settlement and modern vegetation and climates.  We develop a new method, based on Hellinger distances among tree-climate distributions, to quantify the total change and attribute the relative importance of historical vegetation and climate change. We assess whether climate and vegetation changes are compounding, counteracting or confounding. We use this framework to discuss the potential impacts of historical land-use and climate change on inferences about species resilience to climate change, as predicted by species-distribution models based soley on modern distributional data.

@@

Estimates of settlement-era forest composition and tree distribution are based on survey data from the PLS [@white1983history;@schulte2001original;@goring2016preset], aggregated to a 64 km^2^grid across the upper Midwestern United States [@goring2016preset]. We briefly review these methods. PLS data originally were collected on a 1.61 x 1.61 km (1 x 1 mi) grid across the upper Midwest from 1830 to 1910 [@schulte2001original;@white1983history]. At each survey point surveyors noted the closest two to four trees and recorded their common name, distances from survey points, diameters at breast height, and azimuth to trees. Goring et al. [-@goring2016preset] cleaned and aggregated this data to an 8x8 km grid with an average of 61 survey points (120 trees) per cell. This process included data cleaning, taxonomic standardization, and the application of spatially varying correction factors [@goring2016preset]. Using this data, we produce presence-absence distributions for each taxon, across the upper Midwest for 15 common tree genera. A few common names, such as "Ironwood" can not be clearly resolved to a single tree genera and so we combine *Ostrya* and *Carpinus*, and *Thuja* and *Juniperus*. We define 'Presence' as at least one recorded tree in any 8x8 km grid cell and use the term 'range' as the extent in climate space of all cells with recorded presence for a specific taxon.

Modern forest data are from the US FIA program, which provides detailed inventories of contemporary tree distributions, for all regions with >10% forest cover [@woudenberg2010forest]. The FIA uses a nationally standardized sampling procedure that has evolved over time.  In the Phase 2/3 system, beginning in 1998, each plot consists of four, 7.32m radius, circular sub-plots which are sampled intensively [@woudenberg2010forest].

Comparisons between historical and modern vegetation must address differences in sampling design between the PLS and FIA datasets [@goring2016preset;@kronenfeld2014validating;@kronenfeld2010influence]. Here, the primary solution for maximizing comparability is via aggregation, both spatially, by aggregating records from both datasets to a common scale (8x8 km), and taxonomically by aggregating taxa to the level of genera. Nonetheless, differences persist.  The FIA contains fewer plots per 8 km cell, with more trees per plot than the PLS data and FIA plots are sited within forested or partially forested landscapes [@woudenberg2010forest]. while the PLS sampling is uniform. Scattered trees may be present in regions without permanent or temporary FIA plots, complicating the assignments of 'absence' within the FIA. However, the practice of using FIA data as a surrogate for tree distribution data is widespread [@iverson2013tree;@iverson1998predicting], so these complications are general. In prior comparative analyses of the PLS and FIA data, Goring et al. [-@goring2016preset] tested the effect of differential sampling design on the mapped patterns of novel and disappearing forests, and found little effect.

Given the differences in sampling design, FIA data should be better at detecting species presence at the level of individual plots, while detectability at the level of 8km grid cell (the analysis unit of this study) should be similar between FIA and PLS data or higher in the latter.  Within grid cells, there is a tradeoff between the intensive sampling within FIA plots versus more PLS points per grid cell and more uniform sampling of landscape heterogeneity.  The PLS data may be better at detecting tree presence in low-density regions, due to the variable radius of PLS plots.  Both the FIA and PLS datasets both have good coverage in the upper Midwest, with   8km grid cells with at least one FIA plot and XXX  grid cells with at least one PLS point.  In mapping tree distributions in the upper Midwest we consider three classes: "gain", where a tree species was present within a grid cell in the FIA, but absent in the PLS; "loss", where a tree was present in a grid cell during the PLS but absent in the FIA; and "continuous presence", where a tree was present within a grid cell in both FIA and PLS eras.  We present gain and loss estimates for all grid cells (Supplementary Table 1) and for the subset of grid cells with both PLS and FIA data (Supplementary Table  2). For all species, the range within the upper Midwest incompletly samples species ranges, however, the region contains several major ecotones and the southern and western range limit for many tree species [@curtis1959vegetation] so this region comprises an important boundary in geographic and climate space.

@@

Three periods were selected for calculation of climate normals: 1895 to 1919 for the historical period and two alternative for the modern period; 1950 - 1979 and 1985 - 2014.  There is an unavoidable temporal mismatch between the historical climate data and the PLS survey period, on the order of 50 to 70 years; few meteorological records exist in this region prior to 1895 CE, and no gridded dataset exists at high spatio-temporal resolution for this time period for all climate variables.  The effects of temporal mismatch should be minor given that Northern Hemisphere temperature trends were generally small during the 19^th^ century [@mann2009global].  To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we compare the instrumental record against dendroclimatic reconstructions of the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`) and to early temperature observations from US military forts and observer stations, which extend to the 1830s [@baker1985minnesota;@burnette2010daily].

The use of alternative modern climate normals reflects two reasonable choices for whether to pick climate normals corresponding to time of sampling versus time of establishment.  The 1985-2014 normal encompasses the period of sampling (1998 - 2015) for the FIA data.  The 1950  - 1979 normal encompases the time of establishment of FIA trees, which have an average age of 50 years in this region based on the raw stand data.  We report results for the 1985 - 2014 normals, with the 1950-1980 results show in in Supplmentary information.

@@

![](figures/figureFour_panels.tiff)

@@

The total observed shift ($d_{tot}$) between a historical and modern PDF is the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ (`r fig_nums("fourPanel", display = "cite")`a: top left to bottom right). The conditions $V_{M}C_{H}$ and $V_{H}C_{M}$ are synthetic, generated by superimposing historical climate patterns on modern vegetation distributions and *vice versa*.  The effect of climate change on shifting PDFs is $d_{c}$, the difference between PDFs using the early climate normals ($C_{H}$) and modern climate normals ($C_{M}$) overlaid on PLS vegetation (e.g., $V_{H}C_{H}$ versus $V_{H}C_{M}$; `r fig_nums("fourPanel", display = "cite")`a; left column). The effect of vegetation change on shifting PDFs is similarly described as $d_{v}$, the Hellinger distance between $V_{H}C_{H}$ and $V_{M}C_{H}$ (top row of Figure 2a). Because Euro-American land use is the dominant driver of vegetation change in the upper Midwest over the past two centuries (Introduction), we often attribute shifts recorded by $d_{v}$ to the land use change associated with Euro-American settlement.

We develop an attribution index, ($\Delta_{d} = d_{c} - d_{v}$), such that $\Delta_{d} < 0$ indicates stronger influence of historical land use on shifting climate distributions, while $\Delta_{d}> 0$ indicates a stronger influence of historical climate change.

To assess whether the effects of historical land use and climatic changes are compounding, counteracting or confounding we perform t-tests, using the conservative Bonferroni correction to account for multiple tests and adjustments for spatial autocorrelation (Supplementary Information).  For each taxon and climate variable pair we test whether $V_{H}C_{H}$ and $V_{H}C_{M}$ (attribution to climate change), and $V_{H}C_{H}$ and $V_{M}C_{H}$ (attribution to vegetation change) distributions are significantly different.   If both changes are significant and of the same sign, then the shift is compounding.  If both are significant and in opposite directions then the shift is counteracting.  If climate change is significant, but land use is not significant, then the change is defined as confounding.  <!-- We should define the other as confounding as well, since it is, it makes it seem like climate has changed. -->

@@

For most tree genera, historical range losses dominate over gains and are spatially structured (Supplemental Table 1, Figure 3). Range losses in the upper Midwest are concentrated in the south and west, along the prairie-savanna-forest ecotone (Figure 3). For *Ulmus* and *Larix*, however, losses occur throughout the range.  Losses are highest for *Ostrya*/*Carpinus* and *Larix* when assessed as a proportion of total range (Supplemental Table 1), and range losses are high generally, with an average of `r round(mean(loss_table$Loss), 1)`% across all taxa, *Larix*, *Ulmus* and *Betula* have the largest absolute losses in range within the upper Midwest.  Gains, conversely, account for no more than `r round(max(loss_table$Gain),1)`% of any taxon's historical range, with the highest gains observed for *Populus* (`r fig_nums("gain_loss", display="cite")`) and average `r round(mean(loss_table$Gain),1)`% across all taxa.   Spatial patterns of range gains vary widely among tree genera, and often show less spatial bias than losses (Figure 3).  *Picea*, *Larix*, *Abies*, *Tsuga*, *Fagus*, *Ostrya*/*Carpinus*, *Tilia*, and *Fraxinus*, for example, all show a dispersed pattern of gains in few and widely scattered cells, suggesting that these apparent range gains are mainly caused by sampling uncertainty inherent in local-scale plots and stand heterogeneity, rather than systematic range gains.  Conversely, gains for *Populus* and *Acer*, are concentrated in northern Minnesota, Wisconsin, and Michigan, suggesting range infilling, while *Quercus* gains are consistent with some range extension northwards.  *Pinus*, *Juniperus*/*Thuja*, and *Ulmus* range gains are limited but primarily in the southern portion of their historical range, or beyond it.

```{r table1, results="asis", echo=FALSE, eval=TRUE}

sci_names <- c("Larix", "Pinus", "Picea", "Abies", "Tsuga",
               "Thuja/Juniperus", "Populus", "Acer", "Betula", "Fagus", 
               "Ostrya/Carpinus", "Tilia", "Fraxinus", "Ulmus", "Quercus")

rownames(loss_table) <- paste0("*", sci_names, "*")
```

```{r densities, warning=FALSE, message=FALSE, echo=FALSE, dpi=150}
source("R/resampling_climate.R")
vegclim_table <- resample_climate(climate, agg_dens, pls_data)

@@

`r fig_nums("boxplots", "Changes in the realized climate distributions for tree genera in the upper Midwestern United States. In the box-and-whisker plots, the central line indicates the median, box margins indicate 25^th^ and 75^th^ percentiles, whisker tips indicate 5^th^ and 95^th^ percentiles, and dots indicate outliers beyond the 95^th^ percentile. These shifts in realized climate distributions are produced by the joint effects of historical climate change and land-use change.")`

The joint effects of historical climate and vegetation change on tree-climate distributions are illustrated for *Quercus* (Figure 2).  For *Quercus* we see tree-climate PDFs when vegetation is held constant (from solid to dashed lines), regardless of whether modern or pre-settlement vegetation distributions are used, indicating that historical climate change is primarily responsible for shifting the distribution of *Quercus* in climate space. **DOUBLE CHECK THIS**

@@

Both the total shift in realized tree-climate distributions and the attribution of these changes to historical vegetation or climate change vary  among taxa and among climate variables (`r fig_nums("hellPlot", display = "cite")`). Shifts in realized climate distributions over the last 100 to 150 years are largest for *Ulmus* ($T_{min}$, $T_{max}$), *Larix* ($T_{max}$), *Tsuga* ($T_{max}$ and $P_{ann}$), and *Fagus* ($P_{ann}$).   Attribution analyses based on $d$ (`r fig_nums("hellPlot", display = "cite")`) reveal that while the imprint of historical vegetation change on shifting climatic niches is important for many taxa, the effect varies across taxa. The climatic signal is strongest for $P_{ann}$, with regional changes in $P_{ann}$ being largely responsible for shifting climate niches, while only *Populus* and *Picea* have seen climate niche shifts **that are more due** to land-use than climate shifts. The greatest land use signal across taxa is found in $T_{max}$ (`r fig_nums("hellPlot", display = "cite")`), with effects particularly strong for *Larix*, *Ulmus*, and *Fagus* (all taxa lie below the equality line for $d_{c} - d_{v}$). The large shift in the distribution of *Tsuga* relative to $T_{max}$, conversely, appears to be mainly attributable to historical climate change. For $T_{min}$ (`r fig_nums("hellPlot", display = "cite")`), shifting distributions of *Ulmus* and *Thuja*/*Juniperus* can be mainly attributed to land use change, while *Larix* and *Betula* have large shifts that are equally attributable to historical climate change and land use. The niche shifts for *Quercus*, *Fraxinus*, *Picea*, *Acer*, and *Tilia* are attributed to changing climate.

>>>>

* Anthropocene, climate change, climate disequillibrium, forest inventory and analysis (FIA), fundamental niche, historical ecology, land-use, niche shift, Public Land Survey System, realized niche

@@

Contemporary forest inventory data are widely used to understand environmental controls on tree species distributions and to construct models to project forest responses to climate change, but the stability and representativeness of contemporary tree-climate distributions are poorly understood. We show that tree-climate relationships for 15 tree genera in the upper Midwestern US have significantly altered over the last two centuries due to historical land-use and climate change. Realized niches have shifted towards higher minimum temperatures, cooler maximum temperatures, and higher rainfall. A new attribution method implicates both historical climate change and land-use in these shifts, with the relative importance varying among genera and climate variables. Most climate/land-use interactions are compounding, in which historical land-use reinforces climatic shifts in species distributions toward cooler or wetter distributions. The prevalence of compounding interactions imply that contemporary-based models of species distributions may underestimate species resilience to climate change.

@@

Increasingly, ecologists are turning to the paleohistorical record to test the assumptions underlying ecological models and improve their predictive capacity [@moritz2013future;@maguire2015modeling]. Projections relying only on contemporary observational data may overlook plasticity in species-climate responses [@maiorano2013building], incompletely delimit species fundamental niches [@jackson2000responses], miss responses to past no-analog climates [@veloz2012no], or fail to detect effects of anthropogenic land-use, and other historical legacies [@svenning2015influence]. Paleoecological records show that forest responses lag abrupt climate change by a few years to centuries [@williamsinpress;@webb1986vegetation;@ordonez2013realized;@svenning2015influence], suggesting that tree distributions have yet to fully respond to recent climate changes. In principle, mechanistic dynamic vegetation models [*e.g.*, @leiblein2016modelling] can represent the physiological, demographic, and dispersal processes that cause tree populations to lag climate change.  In practice, however, the parameterization of these models is poorly constrained for processes operating at long time scales [@moorcroft2006close;@matthes2016benchmarking].

Historical land-use has substantially affected species distributions and vegetation structure [*e.g.*, @gehrig2007tree] and may cause species distributions to incompletely fill available climates, or otherwise be at disequilibrium [@early2014climatic;@sarmento2013impacts;@ivory2016niche]. Anthropogenic land-use is structured and is often biased within climate space, hence, species distributions and protected natural areas often represent a biased selection from the available environmental space [@scott2001nature]. Biasing can confound studies that use contemporary distributional data to understand the environmental controls on species distributions and predict species responses to future climate scenarios [@pyke2004habitat].

Interactions between climate and land-use change can be classified into three types: compounding, confounding and counteracting [modified from @pyke2004habitat]. Compounding effects occur when climatic and land-use changes cause species distributions to shift in the same direction along an environmental gradient.  For example, warming has been compounded by land-use conversion at the lowest elevations in the northern Sierra Nevada Mountains of California, both causing upslope distributional shifts for butterfly species [@forister2010compounded]. Confounding effects may cause little overall change in species distributions, with land-use pressure occurring across the species' environmental gradient. For example, land-use conversion has excluded Garry oak (*Quercus garryana*) from the deep-soil sites it formerly occupied [@pellatt2014environmental], but land-use has not systematically shifted Garry oak distributions in climate space.  These confounding effects may hinder detection of climatic forcing of historical species distributions. Counteracting effects occur when climate and land-use change act in opposite directions, narrowing the available geographic and climatic space available for species. For example, bird species in the Sierra Nevadas are expected to move downslope due to increasing precipitation [@tingley2012push], but this downslope migration is limited by land-use conversion in the lowlands.

@@

`r fig_nums("land_use_clim", "*(a-e) Climate change in the upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions with a LOESS smooth superimposed for illustration. Modern climate (normals indicated by rectangles).  PDSI shows strong coherence with $P_{ann}$ in this region. (f) Patterns of land-use change in the upper Midwestern United States, with the 'Tension Zone', differentiating sub-boreal from southern broadleafed forests superimposed. land-use data from the NLCD [@jin2013comprehensive].*")`

Tree distributions and community composition in the upper Midwestern US have been heavily influenced by land-use conversion associated with Euro-American Settlement in the 19^th^ and 20^th^ Centuries [@goring2016preset;@schulte2007homogenization;@hanberry2012comparison;@rhemtulla2009historical], outbreaks of exotic pathogens [@barnes1979stem], climate changes accompanying the Little Ice Age [@hotchkiss2007response] and current anthropogenic warming [@wicci2011climate]. Modern forests show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System (PLS) prior to widespread land-use conversion in the late 1800s [@goring2016preset;@schulte2007homogenization]. Rhemtulla et al. [-@rhemtulla2009historical] estimate that *Pinus strobus* and *Tsuga canadensis* now occupy only 4% of their original coverage in Wisconsin. Wildland ecosystems have been converted to agricultural and urban land-use [@rhemtulla2009historical], particularly in historical prairie and savanna ecosystems (`r fig_nums("land_use_clim", display = "cite")`f), where almost 50% of southern deciduous savanna was converted to cropland [@rhemtulla2007regional].  In northern forests, ranges of early successional species have expanded and abundances of late-successional species have been reduced due to widespread logging in the 19^th^ and 20^th^ centuries, and ongoing land management [@schulte2007homogenization;@goring2016preset].  Additionally, the formerly dominant *Ulmus americana* has experienced rangewide population collapses caused by the introduction of exotic pathogens [@barnes1979stem], as did *Castanea dentata*, with other tree species currently threatened (*Tsuga canadensis*, *Fraxinus* spp.) [@barnes1979stem;@schlarbaum1998three;@castello1995pathogens].

Here we demonstrate significant shifts in the climatic distributions of tree genera in the upper Midwestern US over the last two centuries due to historical climate and vegetation change, and we diagnose the relative influence of historical climate change and, land-use changes and pathogen outbreaks (for *Ulmus*) on these shifts.  We use gridded historical and contemporary climate observations (PRISM-LT) and estimates of pre-settlement and modern forest vegetation [@goring2016preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwest.  We construct four observed and hypothetical sets of tree-climate relationships, using a 2x2 factorial of pre-settlement and modern vegetation and climates.  We develop a new method, based on Hellinger distances among tree-climate distributions, to quantify the total change and attribute the relative importance of historical vegetation and climate change. We assess whether climate and vegetation changes are compounding, counteracting or confounding. We use this framework to discuss the potential impacts of historical land-use and climate change on inferences about species resilience to climate change, as predicted by species-distribution models based soley on modern distributional data.

@@

Estimates of settlement-era forest composition and tree distribution are based on survey data from the PLS [@white1983history;@schulte2001original;@goring2016preset], aggregated to a 64 km^2^grid across the upper Midwestern United States [@goring2016preset]. We briefly review these methods. PLS data originally were collected on a 1.61 x 1.61 km (1 x 1 mi) grid across the upper Midwest from 1830 to 1910 [@schulte2001original;@white1983history]. At each survey point surveyors noted the closest two to four trees and recorded their common name, distances from survey points, diameters at breast height, and azimuth to trees. Goring et al. [-@goring2016preset] aggregated this data to an 8x8 km grid with an average of 61 survey points (120 trees) per cell. This process included data cleaning, taxonomic standardization, and the application of spatially varying correction factors [@goring2016preset]. Using this data, we produce presence-absence distributions for each taxon, across the upper Midwest for 15 common tree genera. A few common names, such as "Ironwood" can not be clearly resolved to a single tree genera and so we combine *Ostrya* and *Carpinus*, and *Thuja* and *Juniperus*. We define 'presence' as at least one recorded tree in any 8x8 km grid cell and use the term 'range' as the extent in climate space of all cells with recorded presence for a specific taxon.

Modern forest data are from the US Forest Inventory and Analysis (FIA) program, which provides detailed inventories of contemporary tree distributions, for all regions with >10% forest cover [@woudenberg2010forest]. The FIA uses a nationally standardized sampling procedure; in the Phase 2/3 system, beginning in 1998, each plot consists of four, 7.32m radius, circular sub-plots which are sampled intensively [@woudenberg2010forest].

Comparisons between historical and modern vegetation must address differences in sampling design between the PLS and FIA datasets [@goring2016preset;@kronenfeld2014validating;@kronenfeld2010influence]. Here, the primary solution for maximizing comparability is via aggregation, both spatially, by aggregating records from both datasets to a common scale (8x8 km), and taxonomically by aggregating taxa to the level of genera. Nonetheless, differences persist.  The FIA contains fewer plots per 8 km cell, with more trees per plot than the PLS data, and FIA plots are sited within forested or partially forested landscapes. while the PLS sampling is uniform. Scattered trees may be present in regions without permanent or temporary FIA plots, complicating the assignments of 'absence' within the FIA. However, the practice of using FIA data as a surrogate for tree distribution data is widespread [@iverson2013tree;@iverson1998predicting], so these complications are general. In prior comparative analyses of the PLS and FIA data, Goring et al. [-@goring2016preset] tested the effect of differential sampling design on the mapped patterns of novel and disappearing forests, and found little effect.

Given the differences in sampling design, FIA data should be better at detecting species presence at the level of individual plots, while detectability at the level of 8km grid cell (the analysis unit of this study) should be similar between FIA and PLS data or higher in the latter.  Within grid cells, there is a tradeoff between the intensive sampling within FIA plots versus more PLS points per grid cell and more uniform sampling of landscape heterogeneity.  The PLS data may be better at detecting tree presence in low-density regions, due to the variable radius of PLS plots.  Both the FIA and PLS datasets both have good coverage in the upper Midwest, with 5172 8km grid cells with at least one FIA plot and 7376 grid cells with at least one PLS point.  In mapping tree distributions in the upper Midwest we consider three classes: "gain", where a tree species was present within a grid cell in the FIA, but absent in the PLS; "loss", where a tree was present in a grid cell during the PLS but absent in the FIA; and "continuous presence", where a tree was present within a grid cell in both FIA and PLS eras.  We present gain and loss estimates for all grid cells (Supplementary Table 1) and for the subset of grid cells with both PLS and FIA data (Supplementary Table  2). For all species, the range within the upper Midwest incompletly samples species ranges, however, the region contains several major ecotones and the southern and western range limit for many tree species [@curtis1959vegetation] so this region comprises an important boundary in geographic and climate space.

@@

Three periods were selected for calculation of climate normals: 1895 to 1919 for the historical period and two alternative for the modern period; 1950 - 1979 representing the average period of establishment for stand in the upper Midwest and 1985 - 2014 covering the period of sampling for the FIA data. We report results for the 1985 - 2014 normals, with the 1950-1980 results shown in Supplmentary Information. There is an unavoidable temporal mismatch between the historical climate data and the PLS survey period, on the order of 50 to 70 years; few meteorological records exist in this region prior to 1895 CE, and no gridded dataset exists at high spatio-temporal resolution for this time period for all climate variables.  The effects of temporal mismatch should be minor given that Northern Hemisphere temperature trends were generally small during the 19^th^ century [@mann2009global].  To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we compare the instrumental record against dendroclimatic reconstructions of the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`) and to early temperature observations from US military forts and observer stations, which extend to the 1830s [@baker1985minnesota;@burnette2010daily].

@@

![](Final_Figures/figure_two.tiff)

@@

The total observed shift ($d_{tot}$) between a historical and modern PDF is the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ (`r fig_nums("fourPanel", display = "cite")`a: top left to bottom right). The conditions $V_{M}C_{H}$ and $V_{H}C_{M}$ are synthetic, generated by superimposing historical climate patterns on modern vegetation distributions and *vice versa*.  The effect of climate change on shifting PDFs ($d_{c}$) is the difference between PDFs using the early climate normals ($C_{H}$) and modern climate normals ($C_{M}$) overlaid on PLS vegetation (e.g., $V_{H}C_{H}$ versus $V_{H}C_{M}$; `r fig_nums("fourPanel", display = "cite")`a; left column). The effect of vegetation change on shifting PDFs is $d_{v}$, the Hellinger distance between $V_{H}C_{H}$ and $V_{M}C_{H}$ (top row of Figure 2a). Because Euro-American land-use is the dominant driver of vegetation change in the upper Midwest over the past two centuries (Introduction), we often attribute shifts recorded by $d_{v}$ to the land-use change associated with Euro-American settlement.

We develop an attribution index, ($\Delta_{d} = d_{c} - d_{v}$), such that $\Delta_{d} < 0$ indicates stronger influence of historical land-use on shifting climate distributions, while $\Delta_{d}> 0$ indicates a stronger influence of historical climate change.

To assess whether the effects of historical land-use and climatic changes are compounding, counteracting or confounding we perform t-tests, using the conservative Bonferroni correction to account for multiple tests and adjustments for spatial autocorrelation (Supplementary Information).  For each taxon and climate variable pair we test whether $V_{H}C_{H}$ and $V_{H}C_{M}$ (attribution to climate change), and $V_{H}C_{H}$ and $V_{M}C_{H}$ (attribution to vegetation change) distributions are significantly different.   If both changes are significant and of the same sign, then the shift is compounding.  If both are significant and in opposite directions then the shift is counteracting.  If climate change is significant, but land-use is not significant, then the change is defined as confounding.  <!-- We should define the other as confounding as well, since it is, it makes it seem like climate has changed. -->

@@

All tree genera show higher losses than gains (Supplemental Table 1, Figure 3). Range losses in the upper Midwest are concentrated in the south and west, along the prairie-savanna-forest ecotone (Figure 3). For *Ulmus* and *Larix*, however, losses occur throughout the range.  Losses are highest for *Ostrya*/*Carpinus* and *Larix* when assessed as a proportion of total range (Supplemental Table 1), and range losses are high generally, with an average of `r round(mean(loss_table$Loss), 1)`% across all taxa, *Larix*, *Ulmus* and *Betula* have the largest absolute losses in range within the upper Midwest.  Gains, conversely, account for no more than `r round(max(loss_table$Gain),1)`% of any taxon's historical range, with the highest gains observed for *Populus* (`r fig_nums("gain_loss", display="cite")`) and average `r round(mean(loss_table$Gain),1)`% across all taxa.  Spatial patterns of gains vary widely among tree genera, and often show less spatial bias than losses (Figure 3).  *Picea*, *Larix*, *Abies*, *Tsuga*, *Fagus*, *Ostrya*/*Carpinus*, *Tilia*, and *Fraxinus*, for example, all show a dispersed pattern of gains in few and widely scattered cells, suggesting that these apparent range gains are mainly caused by sampling uncertainty inherent in local-scale plot and stand heterogeneity, rather than systematic range gains.  Conversely, gains for *Populus* and *Acer*, are concentrated in northern Minnesota, Wisconsin, and Michigan, suggesting range infilling, while *Quercus* gains are consistent with some range extension northwards.  *Pinus*, *Juniperus*/*Thuja*, and *Ulmus* range gains are limited but primarily in the southern portion of their historical range, or beyond it.

```{r densities, warning=FALSE, message=FALSE, echo=FALSE, dpi=150}

sci_names <- c("Larix", "Pinus", "Picea", "Abies", "Tsuga",
               "Thuja/Juniperus", "Populus", "Acer", "Betula", "Fagus", 
               "Ostrya/Carpinus", "Tilia", "Fraxinus", "Ulmus", "Quercus")
rownames(loss_table) <- paste0("*", sci_names, "*")

source("R/resampling_climate.R")
vegclim_table <- resample_climate(climate, agg_dens, pls_data)

@@

`r fig_nums("boxplots", "Changes in the realized climate distributions for tree genera in the upper Midwestern United States produced by the joint effects of historical climate change and land-use change. The central line indicates the median, box margins indicate 25^th^ and 75^th^ percentiles, whisker tips indicate 5^th^ and 95^th^ percentiles, and dots indicate outliers beyond the 95^th^ percentile.")`

@@

Both the total shift in realized tree-climate distributions and the attribution of these changes to historical vegetation or climate change vary among taxa and among climate variables (`r fig_nums("hellPlot", display = "cite")`).  The joint effects of historical climate and vegetation change on tree-climate distributions are illustrated for *Quercus* (`r fig_nums("fourPanel", display = "cite")`).  For *Quercus* we see changes due to both both climate (`r fig_nums("fourPanel", display = "cite")`b to d) and due to vegetation change (`r fig_nums("fourPanel", display = "cite")`b to c), but a stronger apparent shift for historical climate change, reflected in the atrtibution index (`r fig_nums("hellPlot", display = "cite")`a).

### 2016-12-02 SupplementaryInformation.Rmd
Revisions ready to pass back to @iceageecologist.
Generally minor text changes, some changes to the `bib` file.  Filled in values,

Cleaned up the first `if` statement in `checker` in `confounding_table`.

Removed sets of files as well to clean up the repo.

If a 1950-1980 climate normal is used instead of 1990-2014 climate normal, then between 1895-1924 and 1950-1980, the regional climate changes are as follows: T_min increased XXoC between 1895-1924 and 1950-1980, T_max changed by XXoC, while T_diff changed by XXoC

```{r}

>>>>

If a 1950-1980 climate normal is used instead of 1990-2014 climate normal, then between 1895-1924 and 1950-1980, the regional climate changes are as follows: T_min increased 0.5^o^C between 1895-1924 and 1950-1980, T_max changed by 0.35^o^C, while T_diff changed by -0.15^o^C
